Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ESVs and ASVs

### 2021-01-20 submission/manuscript.Rmd
Generate lumping and splitting rate figure
Generate ROC curve figure
Create figure captions and integrate numbers into manuscript

To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB; CITATION). Among the `r n_distinct(metadata$species)` species represented in the rrnDB there were `r nrow(metadata)` genomes. The median number of rrn operson per species ranged between `r single_copy$median` (e.g. *`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the rrn operon. As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased (**FIGURE**). On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v4`, `r rate_v34`, and `r rate_v45` variants when considering the V4, V3-V4, and V4-V5 regions of the gene, respectively. Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**). For example, *`r single_copy$species`* generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene. Similarly, a *E. coli* genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome. Across the `r ecoli_copy$N` *E. coli* genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene. These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.

@@

A method to avoid splitting a single genome into multiple units of inference is to cluster 16S rRNA gene sequences together that are similar. However, this also increases the risk of lumping together genes from different species that are similar to each other. Therefore, I assessed the impact of the threshold used to define clusters of 16S rRNA genes on the propensity to split a genome apart or to lump species together. For full length 16S rRNA gene sequences, I found that at a threshold of `r format_pct(threshold_v19)`%, 95% of the species with 7 copies of the rrn operon would be represented by a single OTU. Similarly, thresholds of `r format_pct(threshold_v4)`, `r format_pct(threshold_v34)`, and `r format_pct(threshold_v45)`% were observed for the V4, V3-V4, and V4-V5 regions, respectively. However, at these thresholds, multiple species could be represented by the same OTU. At the highest level of resolution, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively. At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU when considering full length sequences and `r format_pct(lump3_v4)`, `r format_pct(lump3_v34)`, and `r format_pct(lump3_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively. Given the risk of splitting a genome into multiple OTUs is more biologically problematic than lumping species together, larger thresholds are advisable.

@@

To provide a more nuanced approach to selecting a threshold, it would be useful to to quantify the sensitivity and specificity of characterizing bacterial species using OTUs defined at different thresholds. I created confusion matrices for multiple regions of the 16S rRNA gene: true positives were those cases where two ASVs were joined in the same OTU and the same species; true negatives were those cases where two ASVs from different OTUs came from different species; false positives were those ASVs that joined the same OTU, but were from different species; and false negatives were those ASVs that joined different OTUs, but were from the same species. By calculating the sensitivity and specificity for each threshold and each region of the 16S rRNA gene, I was able to constuct a receiver operator characteristic curve (ROC). Because the ROC curve represents a range of possible thresholds and sensitivities and specificities, I used two metrics to select the best threshold for defining an OTU. First, I identified the thresholds where the sensitivity and specificity were most similar to each other. For this criterion, the best distance thresholds were `r format_pct(balance_v19)`% (V1-V9), `r format_pct(balance_v4)`% (V4), `r format_pct(balance_v34)`% (V3-V4), and `r format_pct(balance_v45)`% (V4-V5). Second, I identified the distance threshold that resulted in the point on the ROC curve that was closest to perfect classification. For this criterion, the best distance thresholds were `r format_pct(distance_v19)`% (V1-V9), `r format_pct(distance_v4)`% (V4), `r format_pct(distance_v34)`% (V3-V4), and `r format_pct(distance_v45)`% (V4-V5). Surprisingly, these analyses revealed that thresholds near 3% distance balance the risks of splitting genomes into separate OTUs and lumping species into the same OTU.

@@

**Figure 1. ESV rate increases as the number of genomes sampled per species increases**

>>>>

To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB; CITATION). Among the `r n_distinct(metadata$species)` species represented in the rrnDB there were `r nrow(metadata)` genomes. The median number of rrn operson per species ranged between `r single_copy$median` (e.g. *`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the rrn operon. As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased. On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v4`, `r rate_v34`, and `r rate_v45` variants when considering the V4, V3-V4, and V4-V5 regions of the gene, respectively. Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**). For example, *`r single_copy$species`* generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene. Similarly, a *E. coli* genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome. Across the `r ecoli_copy$N` *E. coli* genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene. These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.

@@

A method to avoid splitting a single genome into multiple units of inference is to cluster 16S rRNA gene sequences together that are similar. However, this also increases the risk of lumping together genes from different species that are similar to each other. Therefore, I assessed the impact of the threshold used to define clusters of 16S rRNA genes on the propensity to split a genome apart or to lump species together. For full length 16S rRNA gene sequences, I found that at a threshold of `r format_pct(threshold_v19)`%, 95% of the species with 7 copies of the rrn operon would be represented by a single OTU. Similarly, thresholds of `r format_pct(threshold_v4)`, `r format_pct(threshold_v34)`, and `r format_pct(threshold_v45)`% were observed for the V4, V3-V4, and V4-V5 regions, respectively. However, at these thresholds, multiple species could be represented by the same OTU (**Figure 2**). At the highest level of resolution, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively. At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU when considering full length sequences and `r format_pct(lump3_v4)`, `r format_pct(lump3_v34)`, and `r format_pct(lump3_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively. Given the risk of splitting a genome into multiple OTUs is more biologically problematic than lumping species together, larger thresholds are advisable.

@@

To provide a more nuanced approach to selecting a threshold, it would be useful to to quantify the sensitivity and specificity of characterizing bacterial species using OTUs defined at different thresholds. I created confusion matrices for multiple regions of the 16S rRNA gene: true positives were those cases where two ASVs were joined in the same OTU and the same species; true negatives were those cases where two ASVs from different OTUs came from different species; false positives were those ASVs that joined the same OTU, but were from different species; and false negatives were those ASVs that joined different OTUs, but were from the same species. By calculating the sensitivity and specificity for each threshold and each region of the 16S rRNA gene, I was able to constuct a receiver operator characteristic curve (ROC; **Figure S1**). Because the ROC curve represents a range of possible thresholds and sensitivities and specificities, I used two metrics to select the best threshold for defining an OTU. First, I identified the thresholds where the sensitivity and specificity were most similar to each other. For this criterion, the best distance thresholds were `r format_pct(balance_v19)`% (V1-V9), `r format_pct(balance_v4)`% (V4), `r format_pct(balance_v34)`% (V3-V4), and `r format_pct(balance_v45)`% (V4-V5). Second, I identified the distance threshold that resulted in the point on the ROC curve that was closest to perfect classification. For this criterion, the best distance thresholds were `r format_pct(distance_v19)`% (V1-V9), `r format_pct(distance_v4)`% (V4), `r format_pct(distance_v34)`% (V3-V4), and `r format_pct(distance_v45)`% (V4-V5). Surprisingly, these analyses revealed that thresholds near 3% distance balance the risks of splitting genomes into separate OTUs and lumping species into the same OTU.

@@

**Figure 1. The ratio of number of distinct ASVs per copy of the *rrn* operon increases for a species as the number of genomes sampled increases.** Each point represents a different species and is shaded to be 80% transparent so that when points overlap they become darker. The blue line represents a smoothed fit through the data.

\includegraphics{../figures/lump_split.pdf}

**Figure 2. As the distance threshold used to define an OTU increases, the fraction of genomes split into separate OTUs decreases while the fraction of species that are merged into the same OTU increases.** These data represent the median fractions for both measurements across 100 randomizations. In each randomization, one genome was sampled from each species.

\includegraphics{../figures/roc_curve.pdf}

```{r}
asv_sens_spec <- sensitivity_specificity %>%
	filter(threshold == 0) %>% 
	mutate(sensitivity = format(sensitivity, digits=3, nsmall=3),
				 specificity = format(specificity, digits=3, nsmall=3),
				 pretty = paste(sensitivity, specificity, sep="/")
				 )
v19 <- asv_sens_spec %>% filter(region == "v19") %>% pull(pretty)
v34 <- asv_sens_spec %>% filter(region == "v34") %>% pull(pretty)
v4 <- asv_sens_spec %>% filter(region == "v4") %>% pull(pretty)
v45 <- asv_sens_spec %>% filter(region == "v45") %>% pull(pretty)

**Figure S1. Distance thresholds larger than 3% provide better sensitivity and specificity when assigning ASVs to OTUs when trying to represent species-level classifications.** The sensitivities and specificities for ASVs were `r v19`, `r v34`, `r v4`, and `r v45` for the V1-V9, V3-V4, V4, V4-V5 regions, respectively. The gray diagonal line represents the position of where the sensitivity and specificity were equal. These data represent the median fractions for sensitivity and specificity across 100 randomizations. In each randomization, one genome was sampled from each species.
